---
title: "Predictors of Religiosity and Spirituality"
author: "Rachel Chen, Topher Hanagan, Jakob Hildebrandt"
date: "2026-04-20"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, include=FALSE}
library(readr)
library(dplyr)
library(tidyverse)
library(GGally)
library(tinytex)
library(broom)
data <- read_csv("Test.csv")
```

# Abstract

The level of one's religiosity and spirituality are often associated with one's other beliefs and habits regarding issues considered related in nature to faith, including political beliefs, family relationships, church attendance and prayer habits, among others. After some preliminary exploratory data analysis to get a feel for the spread of our data, we analyzed the group differences between faith traditions, the relationship between spirituality and religiosity, behavioral and political belief predictors of religiosity and spirituality, finally culminating in the complete models that predict the spirituality and religiosity of individuals based on most available factors. The complete models were able account for more of the variance of the religiosity of an individual, with an $R^2$ of .66, than for the spirituality of an individual, which had an $R^2$ of .43.

# Introduction

The dataset that we're working with comes from the PEW Research Center. The data comes from a survey of the American public about religious affiliation, conducted with the intention of tracking changes in U.S. religious affiliation. The survey was intended to get responses from as many Americans as possible, offering surveys in English and Spanish in all 50 states, with online, paper and phone format options. The data contains information about the religious, political, spiritual and social background of those surveyed. After trimming down our dataset to include only those with useful information and those without all NA values, we were left with 93 columns. Our driving question is twofold: 1) How well can we predict religiosity and spirituality based on factors including church attendance, family relationships, marriage, political opinions, religious affiliation and devotion, etc.? 2) Which factors are most effective in predicting religiosity/spirituality of individuals? Does this vary by religious affiliation? As one of the premier immigration destinations in the world, an analysis of the religious landscape in the United States provides key information on how 

In our dataset, many variables were unusable due to lack of data or relevance to our research question. We narrowed our dataset down to 93 variables, of which the 28 most important were related to religious affiliation, current religion, level of religiosity and spirituality on a scale of 1-5, frequency of church service attendance, belief in God, Heaven and Hell, frequency of prayer, reading of scripture and attendance of religious services growing up, sharing beliefs with others, support of legalization of abortion, the religion of the respondents' family, marital status, support of prayer in public schools, birth decade, race and hispanic status, educational attainment, current school enrollment and employment status, immigration and voting status, political party affiliation and views, income bracket and gender. Many other variables exist and are used in this paper, but these are a sampling of many that are related to the question of religiosity and spirituality.

# Exploratory Data Analysis

First, we turned the current religion variable into a factor so that we can create "bins" of how many people fall into each religious tradition.

```{r include=FALSE}
data <- data |> mutate(CURREL = factor(CURREL,
  levels = c(1000, 10000, 20000, 30000, 40001, 40002, 50000, 60000, 70000, 80000, 90001, 90002, 100000, 900000),
  labels = c("Protestant", "Catholic", "Mormon", "Orthodox", "JW", "Other Christian", "Jewish", "Muslim", "Buddhist", "Hindu", "Other Religion", "Other Faith", "Unaffiliated", "Uninterpretable")
))

data <- data |> mutate(FRMREL = factor(FRMREL,
  levels = c(1000, 10000, 20000, 30000, 40001, 40002, 50000, 60000, 70000, 80000, 90001, 90002, 100000, 900000),
  labels = c("Protestant", "Catholic", "Mormon", "Orthodox", "JW", "Other Christian", "Jewish", "Muslim", "Buddhist", "Hindu", "Other Religion", "Other Faith", "Unaffiliated", "Uninterpretable")
))

data <- data |> mutate(RELTRAD = factor(RELTRAD,
  levels = c(1100,1200,1300, 10000, 20000, 30000, 40001, 40002, 50000, 60000, 70000, 80000, 90001, 90002, 100000, 900000),
  labels = c("Evangelical","Mainline","Blackprot", "Catholic", "Mormon", "Orthodox", "JW", "Other Christian", "Jewish", "Muslim", "Buddhist", "Hindu", "Other Religion", "Other Faith", "Unaffiliated", "Uninterpretable")
))
```

Taking a quick

```{r}
cor(data$SPIRPER, data$RELPER)
```

From there, we were able to look at the relationship between some of the other variables and religion.

First, before trying to plot the data, we noticed that there was a much larger number of those who identified as Catholic, Protestant or Unaffiliated than anything else, so we split the histogram facet_wrap into two to account for this (then split it again to account for the much larger prevalence of Mormon, Jewish and Other Faith than all the other religious identities combined) so that the spread of each religion across the different values would be clear. This splitting up of the religious categories will prove useful throughout the duration of the EDA section.


Looking at the correlation between each specific faith and personal spirituality (1=most spiritual, 4= least spiritual, 5=don't know/refused), we notice a few trends: most faiths seemed to have the largest spike at either 1 or 2. Interestingly, the Unaffiliated category also has the largest spikes around 1 and 2, and not 4. This could be that many people who identified as Unaffiliated, although they did not profess a specific religion, still considered themselves spiritual (spiritual but not religious).

```{r, message=FALSE, echo=FALSE, fig.width=10, fig.height=5}
data$SPIRPER[data$SPIRPER == 99] <- 5

data_subset1 <- data |>
  filter(CURREL %in% c("Protestant", "Catholic", "Unaffiliated"))
data_subset2 <- data |>
  filter(CURREL %in% c("Mormon", "Jewish", "Other Faith"))
data_subset3 <- data |>
  filter(CURREL %in% c("Orthodox", "JW", "Other Christian", "Muslim", "Buddhist", "Hindu", "Other Religion", "Uninterpretable"))

ggplot(data = data_subset1, aes(x = SPIRPER)) +
  geom_histogram() +
  facet_wrap(~CURREL) +
  labs(x = "Spirituality")
ggplot(data = data_subset2, aes(x = SPIRPER)) +
  geom_histogram() +
  facet_wrap(~CURREL) +
  labs(x = "Spirituality")
ggplot(data = data_subset3, aes(x = SPIRPER)) +
  geom_histogram() +
  facet_wrap(~CURREL) +
  labs(x = "Spirituality")
```

The spread for religiosity is shifted more to the higher (less religious) values than spirituality: there were only two religions where the plurality (largest spike) was 1 (most religious).

```{r, message=FALSE, echo=FALSE, fig.width=10, fig.height=5}
data$RELPER[data$RELPER == 99] <- 5

data_subset1 <- data |>
  filter(CURREL %in% c("Protestant", "Catholic", "Unaffiliated"))
data_subset2 <- data |>
  filter(CURREL %in% c("Mormon", "Jewish", "Other Faith"))
data_subset3 <- data |>
  filter(CURREL %in% c("Orthodox", "JW", "Other Christian", "Muslim", "Buddhist", "Hindu", "Other Religion", "Uninterpretable"))

ggplot(data = data_subset1, aes(x = RELPER)) +
  geom_histogram() +
  facet_wrap(~CURREL) +
  labs(x = "Religiosity")
ggplot(data = data_subset2, aes(x = RELPER)) +
  geom_histogram() +
  facet_wrap(~CURREL) +
  labs(x = "Religiosity")
ggplot(data = data_subset3, aes(x = RELPER)) +
  geom_histogram() +
  facet_wrap(~CURREL) +
  labs(x = "Religiosity")
```


# Statistical Analysis

To formally investigate the relationships suggested by the exploratory data analysis, we apply several statistical methods, including analysis of variance (ANOVA), correlation analysis, simple linear regression, and multiple linear regression. These methods allow us to test for group differences, measure associations between variables, and identify key predictors of religiosity.

## Section 1: Group Differences

We begin by examining whether religiosity differs across religious traditions.

Hypothesis Test:

$H_0$: All religious groups have the same mean religiosity.

$H_a$: At least one group's mean is different.

```{r, echo=FALSE}
data_filtered <- subset(data, RELPER != 5)

anova_model <- aov(RELPER ~ CURREL, data = data_filtered)
summary(anova_model)

eda_summary_table <- data_filtered %>%
  group_by(CURREL) %>%
  summarise(
    mean_religiosity = mean(RELPER, na.rm = TRUE),
    sd_religiosity = sd(RELPER, na.rm = TRUE),
    n = n()
  ) %>%
  arrange(mean_religiosity)

knitr::kable(
  eda_summary_table,
  caption = "Summary Statistics by Religious Group",
  col.names = c("Current Religion", "Mean Religiosity", "SD", "Sample Size")
)
```

The ANOVA results indicate that we reject the null hypothesis, as the p-value is less than 0.05. This suggests that religiosity differs significantly across religious traditions.

To further examine these differences, we computed the mean religiosity score for each group. Religiosity is measured on a scale from 1 (very religious) to 4 (not at all religious), excluding responses coded as 5 (don’t know/refused). Because lower values indicate higher religiosity, groups with lower mean scores are more religious on average. The results show that Protestants have lower mean religiosity scores than Catholics and the unaffiliated, indicating higher average religiosity, while the unaffiliated group reports the highest mean scores, indicating the lowest levels of religiosity.

Protestants, Catholics, and the unaffiliated also have the largest sample sizes, making these estimates more reliable. The standard deviations, which range approximately from 0.5 to 1, suggest a moderate level of variation in religiosity within each group.

```{r, echo=FALSE, fig.width=10, fig.height=5}
summary_df <- data_filtered %>%
  group_by(CURREL) %>%
  summarise(
    mean_religiosity = mean(RELPER, na.rm = TRUE),
    sd_religiosity = sd(RELPER, na.rm = TRUE),
    n = n()
  )

ggplot(summary_df, aes(
  x = reorder(CURREL, mean_religiosity),
  y = mean_religiosity
)) +
  geom_col(fill = "steelblue") +
  geom_errorbar(
    aes(
      ymin = mean_religiosity - sd_religiosity,
      ymax = mean_religiosity + sd_religiosity
    ),
    width = 0.2
  ) +
  coord_flip() +
  labs(
    x = "Religion Groups",
    y = "Mean Religiosity Score",
    title = "Mean Religiosity(with Standard Deviation) by Religious Tradition"
  ) +
  theme_minimal()
```

Check Assumptions for ANOVA:

1. Independence

2. Normality within groups

3. Common variance

```{r, warning = FALSE, echo=FALSE, fig.width=10, fig.height=5}
m <- augment(anova_model)

ggplot(m, aes(x = 1:nrow(m), y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Residuals vs Observation Order (Independence Check)",
    x = "Observation Order",
    y = "Residuals"
  ) +
  theme_minimal()

ggplot(data_filtered, aes(sample = RELPER)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~ CURREL) +
  labs(title = "Normality Check Within Groups (Q-Q Plots)") +
  theme_minimal()

ggplot(m, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Constant Variance Check",
    x = "Fitted Values",
    y = "Residuals"
  ) +
  theme_minimal()
```

To assess the assumptions of ANOVA, several diagnostic plots were examined. The residuals versus observation order plot shows no clear pattern, suggesting that the independence assumption is reasonably satisfied. The Q-Q plots across religious groups indicate that residuals are approximately normally distributed within each group, with only minor deviations in the tails, which is expected given the large sample sizes. Finally, the residuals versus fitted values plot shows a relatively constant spread of residuals, suggesting that the assumption of equal variances across groups is reasonably met. Overall, the diagnostic checks indicate that the assumptions required for ANOVA are adequately satisfied for this analysis.

## Section 2: Spirituality vs Religiosity

Next, we examine the relationship between the spirituality and religiosity variables, as the exploratory data analysis suggested that even unaffiliated individuals report relatively high levels of spirituality.

```{r, echo=FALSE}
data_no5 <- subset(data, RELPER != 5 & SPIRPER != 5)
cor(data_no5$RELPER, data_no5$SPIRPER, use = "complete.obs")

model1 <- lm(RELPER ~ SPIRPER, data = data_no5)
summary(model1)
```

A correlation analysis was conducted between religiosity and spirituality, excluding missing responses. The result is 0.562, indicating a moderate positive relationship between the two variables. This suggests that higher levels of spirituality are associated with higher religiosity scores. Given that higher religiosity scores correspond to lower levels of religiosity (1 = very religious, 4 = not at all religious), this indicates that higher spirituality is associated with lower reported religiosity.

A simple linear regression further examined this relationship. The results show that spirituality is a statistically significant predictor of religiosity (p < 0.05). The coefficient indicates that a one-unit increase in spirituality is associated with an increase in religiosity score. The model explains approximately 31.57% ($R^2$ = 0.3157) of the variation in religiosity, suggesting a moderate relationship.

```{r, echo=FALSE, fig.width=10, fig.height=5, message=FALSE}
heat_data <- data_no5 %>%
  count(SPIRPER, RELPER)
library(ggplot2)
ggplot(heat_data, aes(x = SPIRPER, y = RELPER, fill = n)) +
  geom_tile(color = "white") +
  geom_text(aes(label = n), color = "white", size = 5) +
  scale_fill_gradientn(colors = c("green", "goldenrod", "orange", "red")) +
  scale_x_continuous(breaks = 1:4) +
  scale_y_continuous(breaks = 1:4) +
  theme_minimal() +
  labs(
    x = "Spirituality (1 = very spiritual, 4 = not at all spiritual)",
    y = "Religiosity (1 = very religious, 4 = not at all religious)",
    title = "Heatmap of Spirituality vs Religiosity",
    fill = "Count"
  )
```

The heatmap shows a clear positive relationship between spirituality and religiosity. The highest concentration of respondents falls in the moderate range for both variables, particularly at level 2. Additionally, extreme mismatches, such as being very spiritual but not religious, are quite rare, suggesting that these two constructs tend to align closely for most individuals.

We want to check the assumptions for linear model:

```{r, echo=FALSE, warning=FALSE}
m1 <- augment(model1)

ggplot(data = m1, aes(x = .resid)) +
  geom_histogram(aes(y = ..density..), bins = 10) +
  stat_function(
    fun = dnorm,
    args = list(
      mean = mean(m1$.resid),
      sd = sd(m1$.resid)
    ),
    color = "blue",
    linewidth = 1.5
  ) +
  labs(
    title = "Histogram of Residuals",
    x = "Residuals",
    y = "Density"
  ) +
  theme_minimal()

ggplot(model1, aes(sample = .resid)) +
  stat_qq() +
  stat_qq_line() +
  labs(title = "Q-Q Plot of Residuals") +
  theme_minimal()
```

To assess the normality assumption of the linear model, both a histogram of residuals and a Q-Q plot were examined. The histogram shows that the residuals are approximately bell-shaped and centered around zero, with only minor deviations from normality. The Q-Q plot further indicates that most points lie close to the reference line, with slight deviations in the tails. Given the large sample size, these deviations are not concerning, and the normality assumption is considered reasonably satisfied.

## Section 3: Behavioral Predictors

While spirituality captures internal beliefs, religiosity may also be influenced by observable behaviors such as attendance and prayer. We therefore examine whether these behavioral variables are significant predictors of religiosity.

We focused on four variables here, including frequency of religious service attendance, prayer, scripture reading, and sharing of religious views. We test whether these behavioral factors significantly predict religiosity.

Hypothesis Test:

$H_0$: That none of these behaviors are associated with religiosity.

$H_a$: That at least one behavioral variable significantly predicts religiosity.

```{r, echo=FALSE}
no99 <- subset(data, RELPER != 5 & ATTNDPERRLS != 99 & PRAY != 99 & PRAC_B != 99 & PRAC_D != 99)
model2 <- lm(RELPER ~ ATTNDPERRLS + PRAY + PRAC_B + PRAC_D, data = no99)
summary(model2)
```

The results from the full model indicate that attendance, prayer frequency, and scripture reading are statistically significant predictors of religiosity (p < 0.05). Because religiosity is reverse-coded (higher values = lower religiosity), the positive coefficients indicate that more frequent religious behaviors are associated with lower religiosity scores, that is higher actual religiosity. In contrast, sharing beliefs (PRAC_D) is not statistically significant, suggesting that sharing religious views with others does not have a measurable relationship with religiosity in this model.

The model explains approximately 58.56% ($R^2$ = 0.5856) of the variation in religiosity, indicating a relatively strong relationship between behavioral practices and religiosity.

After check the total group model, we focus on Protestant, Catholic, and Unaffiliated groups because they have the largest sample sizes in the dataset, which provides more reliable and stable estimates. While core practices such as prayer and attendance are strong predictors of religiosity in the Muslim subsample, other behaviors such as scripture reading and sharing beliefs are not statistically significant, suggesting that the relationship between behavior and religiosity is selective rather than uniform.

### Protestant Group

```{r, echo=FALSE}
protestant_data <- subset(no99, CURREL == "Protestant")
model_protestant <- lm(RELPER ~ ATTNDPERRLS + PRAY + PRAC_B + PRAC_D,
  data = protestant_data
)
summary(model_protestant)
```

The results from the Protestant model indicate that attendance (ATTNDPERRLS), prayer frequency (PRAY), scripture reading (PRAC_B), and sharing religious views (PRAC_D) are statistically significant predictors of religiosity (p < 0.05). Given the coding of religiosity, where higher values indicate lower levels of religiosity, these results suggest that more frequent religious practices are associated with lower religiosity scores, indicating higher levels of religious engagement.

The model explains approximately 27.34% ($R^2$ = 0.2734) of the variation in religiosity, suggesting a moderate relationship between behavioral practices and religiosity within the Protestant subsample.

### Catholic Group

```{r, echo=FALSE}
catholic_data <- subset(no99, CURREL == "Catholic")
model_catholic <- lm(RELPER ~ ATTNDPERRLS + PRAY + PRAC_B + PRAC_D,
  data = catholic_data
)
summary(model_catholic)
```

The results from the Catholic model indicate that attendance (ATTNDPERRLS), prayer frequency (PRAY), scripture reading (PRAC_B), and sharing religious views (PRAC_D) are statistically significant predictors of religiosity (p < 0.05). Given the coding of religiosity, where higher values indicate lower levels of religiosity, these results suggest that more frequent religious practices are associated with lower religiosity scores, indicating higher levels of religious engagement.

The model explains approximately 47.64% ($R^2$ = 0.4764) of the variation in religiosity, suggesting a relatively strong relationship between behavioral practices and religiosity within the Catholic subsample.

### Muslim Group

```{r, echo=FALSE}
muslim_data <- subset(no99, CURREL == "Muslim")
model_muslim <- lm(RELPER ~ ATTNDPERRLS + PRAY + PRAC_B + PRAC_D,
  data = muslim_data
)
summary(model_muslim)
```

The results from the Muslim model indicate that attendance (ATTNDPERRLS) and prayer frequency (PRAY) are statistically significant predictors of religiosity (p < 0.05). Given the coding of religiosity, where higher values indicate lower levels of religiosity, these results suggest that more frequent religious practices are associated with lower religiosity scores, indicating higher levels of religious engagement. In contrast, scripture reading (PRAC_B) and sharing beliefs (PRAC_D) is not statistically significant, suggesting that these two behaviors does not have a measurable relationship with religiosity in this model.

The model explains approximately 40.63% ($R^2$ = 0.4063) of the variation in religiosity, suggesting a moderate to relatively strong relationship between behavioral practices and religiosity within the Muslim subsample.

### Unaffiliated Group

```{r, echo=FALSE}
unaffil_data <- subset(no99, CURREL == "Unaffiliated")
model_unaffil <- lm(RELPER ~ ATTNDPERRLS + PRAY + PRAC_B + PRAC_D,
  data = unaffil_data
)
summary(model_unaffil)
```

The results from the Unaffiliated model indicate that some behavioral variables are statistically significant predictors of religiosity (p < 0.05). Given the coding of religiosity, where higher values indicate lower levels of religiosity, these results suggest that more frequent religious practices are associated with lower religiosity scores, indicating higher levels of religious engagement.

The model explains approximately 38.84% ($R^2$ = 0.3884) of the variation in religiosity, suggesting a moderate relationship between behavioral practices and religiosity within the Unaffiliated subsample.

```{r, echo=FALSE, warning=FALSE}
models <- list(
  Overall = model2,
  Protestant = model_protestant,
  Catholic = model_catholic,
  Muslim = model_muslim,
  Unaffiliated = model_unaffil
)

coef_df <- bind_rows(lapply(names(models), function(name) {
  tidy(models[[name]], conf.int = TRUE) |> 
    filter(term != "(Intercept)") |> 
    mutate(Group = name)
})) |>
  mutate(term = recode(term,
    ATTNDPERRLS = "Attendance",
    PRAY = "Prayer",
    PRAC_B = "Scripture Reading",
    PRAC_D = "Sharing Beliefs"
  ))

ggplot(coef_df, aes(x = estimate, y = term, color = Group)) +
  geom_point(position = position_dodge(width = 0.5)) +
  geom_errorbar(
  aes(xmin = conf.low, xmax = conf.high),
  position = position_dodge(width = 0.5),
  width = 0.2,
  orientation = "y") +
  labs(
    title = "Behavioral Predictors of Religiosity",
    x = "Effect on Religiosity",
    y = "Behavioral Variables"
  ) +
  theme_minimal()
```

The coefficient plot compares the effects of behavioral predictors on religiosity across different religious groups. Each point represents the estimated effect size, with horizontal bars indicating 95% confidence intervals. Attendance and prayer consistently exhibit the largest effect sizes across all groups, indicating that these core religious practices are the most robust behavioral predictors of religiosity. In contrast, scripture reading has a smaller but still positive effect, while sharing beliefs shows the greatest variability across groups, including weak or insignificant effects in some cases. Overall, the results indicate that while core devotional practices are strong and stable predictors of religiosity across groups, the influence of interpersonal religious expression varies considerably.

We want to check the assumptions for linear model:

```{r, echo=FALSE, warning=FALSE}
m2 <- augment(model2)

ggplot(data = m2, aes(x = .resid)) +
  geom_histogram(aes(y = ..density..), bins = 10) +
  stat_function(
    fun = dnorm,
    args = list(
      mean = mean(m1$.resid),
      sd = sd(m1$.resid)
    ),
    color = "blue",
    linewidth = 1.5
  ) +
  labs(
    title = "Histogram of Residuals",
    x = "Residuals",
    y = "Density"
  ) +
  theme_minimal()

ggplot(model1, aes(sample = .resid)) +
  stat_qq() +
  stat_qq_line() +
  labs(title = "Q-Q Plot of Residuals") +
  theme_minimal()
```

To assess the normality assumption of the linear model, both a histogram of residuals and a Q-Q plot were examined. The histogram shows that the residuals are approximately bell-shaped and centered around zero, with only minor deviations from normality. The Q-Q plot further indicates that most points lie close to the reference line, with slight deviations in the tails. Given the large sample size, these deviations are not concerning, and the normality assumption is considered reasonably satisfied. The residuals of the overall model appear approximately normal based on the histogram and Q-Q plot. While subgroup models were not individually assessed here, the large sample sizes in most groups reduce concerns about minor deviations from normality due to the Central Limit Theorem. However, caution is warranted for smaller subsamples (e.g., Muslim group), where deviations may have a greater impact.

We estimate separate models by religious group to examine whether the relationship between behavioral practices and religiosity varies across traditions.  Prior research suggests that religious expression differs across groups—for example, Protestant traditions often emphasize personal faith expression, while Catholic practice may be more institutionally structured, and Islamic religiosity is often closely tied to prescribed practices.  By estimating subgroup models, we can test whether behavioral predictors have different magnitudes or significance across these contexts.

## Section 4: Political Predictors of Religiosity

```{r, echo=FALSE}
data_clean <- data |>
  filter(
    RELPER    != 99,
    ABRTLGL   != 99,
    GAYMARR   != 99,
    QB2A      != 99,
    CHNG_C    != 99,
    CHNG_A    != 99,
    DIVRELPOP != 99,
    DIVRACPOP != 99,
    QB2C      != 99
  ) |>
  rename(
    Religiosity            = RELPER,
    Abortion_Legality      = ABRTLGL,
    Gay_Marriage_Support   = GAYMARR,
    Homosexuality_Acceptance     = QB2A,
    Transgender_Acceptance = CHNG_C,
    Immigrant_Impact       = CHNG_A,
    Relig_Diversity        = DIVRELPOP,
    Racial_Diversity       = DIVRACPOP,
    Environmental_Regulation       = QB2C
  )

model3 <- lm(Religiosity ~ Abortion_Legality + Gay_Marriage_Support  + Transgender_Acceptance + Immigrant_Impact + Racial_Diversity + Relig_Diversity + Homosexuality_Acceptance + Environmental_Regulation, data = data_clean)
summary(model3)



m1_resids <- augment(model3) %>% 
  mutate(Model = "Religiosity Model")


```
It would appear that removal of the 99s which represent those who did not answer made a substantial difference on the fit of the model. 
Furthermore, support for gay marriage predicts religiosity well, as does support for transgender individuals. However, acceptance of homosexuality is an incredibly poor predictor of religiosity in the body public. 

Alternatively, support for abortion, homosexuality, and gay marriage predicts spirituality but support for transgender individuals does not predict spirituality.

```{r, echo=FALSE}

catholic_pol <- data_clean |> filter(CURREL == "Catholic")
catholic_polly <- lm(Religiosity ~ Abortion_Legality  + Gay_Marriage_Support + Homosexuality_Acceptance + Transgender_Acceptance, data = catholic_pol)
summary(catholic_polly)


protestant_pol <- data_clean |> filter(CURREL == "Protestant")
protestant_polly <- lm(Religiosity ~ Abortion_Legality  + Gay_Marriage_Support + Homosexuality_Acceptance + Transgender_Acceptance, data = protestant_pol)
summary(protestant_polly)

unaffiliated_pol <- data_clean |> filter(CURREL == "Unaffiliated")
unaffiliated_polly <- lm(Religiosity ~ Abortion_Legality  + Gay_Marriage_Support + Homosexuality_Acceptance + Transgender_Acceptance, data = unaffiliated_pol)
summary(unaffiliated_polly)
```


While support for abortion, same sex marriage, and homosexuality is a statistically significant predictor of religiosity in Catholics and people who are unaffiliated, support for these causes is not statistically significant in predicting religiosity in Protestants. Interestingly, greater acceptance of transgender individuals cannot be used to predict religiosity in people who are unaffiliated. 

```{r, echo=FALSE}


catholic_pol2 <- data_clean |> filter(CURREL == "Catholic")
catholic_polly2 <- lm(Religiosity ~ Immigrant_Impact + Relig_Diversity   + Racial_Diversity + Environmental_Regulation, data = catholic_pol)
summary(catholic_polly2)

protestant_pol2 <- data_clean |> filter(CURREL == "Protestant")
protestant_polly2 <- lm(Religiosity ~ Immigrant_Impact + Relig_Diversity + Racial_Diversity + Environmental_Regulation, data = protestant_pol2)
summary(protestant_polly2)

unaffiliated_pol2 <- data_clean |> filter(CURREL == "Unaffiliated")
unaffiliated_polly2 <- lm(Religiosity ~ Abortion_Legality  + Gay_Marriage_Support + Homosexuality_Acceptance + Transgender_Acceptance, data = unaffiliated_pol2)
summary(unaffiliated_polly2)

all_models <- bind_rows(
  tidy(catholic_polly2, conf.int = TRUE) %>% mutate(Group = "Catholic"),
  tidy(protestant_polly2, conf.int = TRUE) %>% mutate(Group = "Protestant"),
  tidy(unaffiliated_polly2, conf.int = TRUE) %>% mutate(Group = "Unaffiliated")
) %>%
  filter(term != "(Intercept)")

ggplot(all_models, aes(x = estimate, y = term, color = Group)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_point(position = position_dodge(width = 0.6), size = 3) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), 
                 position = position_dodge(width = 0.6), height = 0.3) +
  labs(title = "Social & Diversity Predictors of Religiosity",
       subtitle = "Comparing Regression Coefficients (Slopes) across Religious Traditions",
       x = "Estimate (Impact on Religiosity)",
       y = "Predictor Variable",
       color = "Religious Tradition") +
  theme_minimal() +
  scale_color_brewer(palette = "Set1")



```

## Section 5: Comprehensive Model

Finally, using 26 variables we construct a comprehensive model that incorporates behavioral measures, belief variables, and religious affiliation to determine which factors are most important in predicting religiosity.

```{r, echo=FALSE, include=FALSE}

vars <- c("RELPER", "ATTNDPERRLS", "GOD", "HVN", "HLL", 
          "PRAY", "PRAC_D", "PRAC_B", "ABRTLGL", "CHATTEND", 
          "MARITAL", "SCPRY2", "BIRTHDECADE", "HISP", "RACECMB", "EDUCREC", 
          "E2", "EMPLSIT", "USGEN", "REG", "PARTY", "IDEO", "INC_SDT1", "GENDER")

data_new <- data |>
  mutate(across(all_of(vars), ~ na_if(., 99)))


final_model_rel <- lm(RELPER ~ RELTRAD + ATTNDPERRLS + GOD + HVN + HLL + PRAY + PRAC_D + PRAC_B + ABRTLGL  + FRMREL + CHATTEND + MARITAL + SCPRY2 + BIRTHDECADE + HISP + RACECMB + EDUCREC + E2 + EMPLSIT + USGEN + REG + PARTY + IDEO + INC_SDT1 + GENDER, data = data_new)

summary(final_model_rel)
```

With an $R^2$ of 0.6645, we have pretty good confidence that this model can predict religiosity in individuals. An $R^2$ of 0.66 tells us that 66% of variance can be explained by the model and the variables that we presented it with. The $R^2$ for spirituality was quite a bit lower, however, with an $R^2$ of just .4346.

```{r, echo=FALSE, include=FALSE}
final_model_spir <- lm(SPIRPER ~ RELTRAD +  ATTNDPERRLS + GOD + HVN + HLL + PRAY + PRAC_D + PRAC_B + ABRTLGL  + FRMREL + CHATTEND + MARITAL + SCPRY2 + BIRTHDECADE + HISP + RACECMB + EDUCREC + E2 + EMPLSIT + USGEN + REG + PARTY + IDEO + INC_SDT1 + GENDER, data = data_new)

summary(final_model_spir)
```

Checking the assumptions for regression, we see that the residuals for the spirituality model are normally distributed:

```{r, warning=FALSE, echo=FALSE, fig.width=10, fig.height=5}
m2 <- augment(final_model_spir)

ggplot(data = m2, aes(x = .resid)) +
  geom_histogram(aes(y = ..density..), bins = 10) +
  stat_function(
    fun = dnorm,
    args = list(
      mean = mean(m2$.resid),
      sd = sd(m2$.resid)
    ),
    color = "blue",
    linewidth = 1.5
  ) +
  labs(
    title = "Histogram of Residuals",
    x = "Residuals",
    y = "Density"
  ) +
  theme_minimal()

ggplot(final_model_spir, aes(sample = .resid)) +
  stat_qq() +
  stat_qq_line() +
  labs(title = "Q-Q Plot of Residuals") +
  theme_minimal()
```

The same is true of the religiosity model:

```{r, warning=FALSE, echo=FALSE, fig.width=10, fig.height=5}
m3 <- augment(final_model_rel)

ggplot(data = m3, aes(x = .resid)) +
  geom_histogram(aes(y = ..density..), bins = 10) +
  stat_function(
    fun = dnorm,
    args = list(
      mean = mean(m3$.resid),
      sd = sd(m3$.resid)
    ),
    color = "blue",
    linewidth = 1.5
  ) +
  labs(
    title = "Histogram of Residuals",
    x = "Residuals",
    y = "Density"
  ) +
  theme_minimal()

ggplot(final_model_rel, aes(sample = .resid)) +
  stat_qq() +
  stat_qq_line() +
  labs(title = "Q-Q Plot of Residuals") +
  theme_minimal()
```

```{r, echo=FALSE, include=FALSE}
final_model_rel2 <- lm(RELPER ~ RELTRAD + ATTNDPERRLS + GOD + HVN + HLL + PRAY + PRAC_D + PRAC_B + ABRTLGL + CHATTEND + SCPRY2 + EDUCREC + EMPLSIT + PARTY + IDEO + INC_SDT1 + GENDER, data = data_new)

summary(final_model_rel2)
```

```{r, echo=FALSE, include=FALSE}
final_model_spir2 <- lm(SPIRPER ~ RELTRAD + ATTNDPERRLS + GOD + HVN + PRAY + PRAC_D + PRAC_B + FRMREL + CHATTEND + SCPRY2 + BIRTHDECADE + HISP + EDUCREC + EMPLSIT + USGEN + REG + PARTY + IDEO + INC_SDT1 + GENDER, data = data_new)

summary(final_model_spir2)
```


Looking at the p-values for the religiosity model, there were a few variables that were not impactful in predicting religiosity. These included those related to family religion, marital status, birth decade, race and ethnicity, voter registration, current school enrollment status, and immigration status. When these variables were removed, the $R^2$ value decreased to .6622. The spirituality model had just five variables with a high p-value, which interestingly included those related to belief in Hell (but not Heaven or God) and opinions on the legalization of abortion. Other variables with high p-values included marital status, race and ethnicity.


# Conclusions

Across both the full sample and some religious subsamples, behavioral factors—including attendance, prayer frequency, and scripture reading—consistently emerge as significant predictors of religiosity. These findings suggest that religiosity is strongly shaped by observable religious practices, rather than being solely an internal belief system.

However, the strength of these relationships varies across religious traditions. The Catholic and Muslim subsamples show relatively stronger explanatory power, while the Protestant model shows a weaker relationship between behaviors and religiosity. The Unaffiliated group also demonstrates that behavioral practices remain relevant even in the absence of formal religious affiliation.

Overall, the results indicate that religiosity is best understood as a behaviorally structured phenomenon, but the degree to which behavior predicts religiosity differs across religious traditions. This suggests that religious tradition plays a moderating role in shaping how behaviors translate into levels of religiosity.

Similarly, depending on the religious tradition, certain political beliefs seemed to show a moderate correlation with religiosity and spirituality.

For the full models, religiosity was much easier to predict based on the behavioral, political and religious beliefs and factors than spirituality, as the religiosity model had an $R^2$ value of .66 compared to the $R^2$ value of .43 for the spirituality model. One reason that this may be the case is potentially due to religious affiliation (with higher religiosity being already more rare) being associated with a specific set of beliefs that one subscribes to, while for spirituality one can claim to be spiritual without subscribing to a specific set of beliefs; this term is more vague in what it describes than religiosity. Many classify themselves as "spiritual but not religious," a category that denotes vague spiritual feelings without adherence to a specific religion. It is important to note that the analogous "religious but not spiritual" designation is far less common.

# Appendix

## List of Variables

The codebook that was available for download with the dataset gave a description of each variable that included the full question that was asked on the survey. The list of 28 primary variables of interest used in this report are listed in their original form along with their description and potential values below:




Source:
https://www.pewresearch.org/dataset/2023-24-religious-landscape-study-rls-dataset/