---
title: "religion_project"
author: "Jake Hildebrandt"
date: "2026-03-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(readr)
library(dplyr)
#change
data <- read_csv("Test.csv")

data <-data |> select(-REGION, -DENOMREC2, -FAMILY, -PROTFAM, -CURREL2, -FRMREL2,-SPREL, -SPREL2, -SPBORNFINAL,-KIDS4ANDUNDERREC,-KIDS5TO12REC,-KIDS13TO17REC,-CITIZEN,-BRANCH,-BIRTHHALFDECADE,-RESPONDENT_BIRTHREGION,-MOTHER_BIRTHREGION,-FATHER_BIRTHREGION,-YEARSINUSREC,-ORIENTMOD,-STATE,-MSA)
```
This dataset has information about over 36,000 individuals, with columns including information about the state or origin, happiness, questions about attitudes about other religions and immigrants, religious teachings, perceived spirituality/religiosity, marriage status, political leaning, family income, among others. This data comes from the Pew Research Center, a reputable source for surveys on the American population.
Some potential areas we could explore with this data include seeing which variables are most associated with religiosity, 
