---
title: "religion_project"
author: "Rachel Chen, Topher Hanagan, Jakob Hildebrandt"
date: "2026-03-29"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, include=FALSE}
library(readr)
library(dplyr)
library(tidyverse)
# import data
data <- read_csv("Test.csv")
```

This dataset has information about over 36,000 individuals, with columns including information about the state or origin, happiness, questions about attitudes about other religions and immigrants, religious teachings, perceived spirituality/religiosity, marriage status, political leaning, family income, among others. This data comes from the Pew Research Center, a reputable source for surveys on the American population.

### Ideas and Questions:

Some potential areas we could explore with this data include seeing which variables are most associated with religiosity.

1) Test people's faithfulness as predictors or stand-ins. How important is religion in your life? How important is this to you when making decisions about right and wrong?

2) Are you personally a member of a church, synagogue, mosque, temple, or other house of worship?

3) Look at people who consider themselves religious vs actually attend religious services. 

4) Try to find the correlation between consider yourself religious and attendance. Could also use a composite for praying and attendance to as a stand-in for "religiosity"

5) Check how the religion related with family or parents. How parents religion affect your religion? Is father religion or mother religion same as your?

6) How religious based of age(birth)? Try to find out did the state affect how you religious?

7) How marriage affect the religion?

8) How support for abortion vs gay marriage as a predictor for party affiliation?



# Introduction
The dataset that we're working with comes from the PEW Research Center. The survey was intended to get responses from as many Americans as possible, offering surveys in English and Spanish in all 50 states, with online, paper and phone format options. The data contains information about the religious, political, spiritual and social background of those surveyed. The data can be accessed at the following link: https://www.pewresearch.org/dataset/2023-24-religious-landscape-study-rls-dataset/
After trimming down our dataset to include only those with useful information and those without all NA values, we were left with 93 columns. Our driving question is twofold: 1) How well can we predict religiosity and spirituality based on factors including church attendance, family relationships, marriage, political opinions, religious affiliation and devotion, etc.?
2) Which factors are most effective in predicting religiosity/spirituality of individuals? Does this vary by religious affiliation?

# Exploratory Data Analysis

In our dataset, the most important variables are:

RELTRAD (religious affiliation)

CURREL (same as before but collapses evangelical, mainline and black prot into one group)

RELPER (how religious the respondent is, 1-4)

SPIRPER (how spiritual the respondent is, 1-4)

ATTNDPERRLS (how often they attend in person)

GOD (belief in God)

HVN (is there a heaven)

HLL (is there a hell)

PRAY (frequency of prayer in personal life)

PRAC_B (how often do you read scripture outside of worship service)

PRAC_D (how often do you share your beliefs with others)

Political Correlation was also observed. 
Variables included:
ABRTLGL (support for legalization of abortion in all or most cases)
CHNG_A ()
CHNG_B 
DIVRELPOP 
DIVRACPOP
QB2C, 

First, we turn CURREL into a factor so that we can create "bins" of how many people fall into each religious tradition

```{r}
data <- data |> mutate(CURREL = factor(CURREL, levels=c(1000,10000,20000,30000,40001,40002,50000,60000,70000,80000,90001,90002,100000,900000),
                                       labels=c("Protestant","Catholic", "Mormon", "Orthodox", "JW", "Other Christian", "Jewish", "Muslim", "Buddhist", "Hindu", "Other Religion", "Other Faith", "Unaffiliated", "Uninterpretable")))
```

From here, we can look at the relationship between each of the other variables and religion.


First, before trying to plot the data, we noticed that there was a much larger number of those who identified as Catholic, Protestant or Unaffiliated than anything else, so we split the histogram facet_wrap into two to account for this (then split it again to account for the much larger prevalence of Mormon, Jewish and Other Faith than all the other religious identities combined) so that the spread of each religion across the different values would be clear. This splitting up of the religious categories will prove useful throughout the duration of the EDA section.


Looking at the correlation between each specific faith and personal spirituality (1=most spiritual, 4= least spiritual, 5=don't know/refused), we notice a few trends: most faiths seemed to have the largest spike at either 1 or 2. Interestingly, the Unaffiliated category also has the largest spikes around 1 and 2, and not 4. This could be that many people who identified as Unaffiliated, although they did not profess a specific religion, still considered themselves spiritual (spiritual but not religious).

```{r, message=FALSE}
data$SPIRPER[data$SPIRPER == 99] <- 5

data_subset1 <- data |>
  filter(CURREL %in% c("Protestant", "Catholic", "Unaffiliated"))
data_subset2 <- data |>
  filter(CURREL %in% c("Mormon", "Jewish", "Other Faith"))
data_subset3 <- data |>
  filter(CURREL %in% c("Orthodox", "JW", "Other Christian", "Muslim", "Buddhist", "Hindu", "Other Religion",  "Uninterpretable"))

ggplot(data = data_subset1, aes(x=SPIRPER))+
  geom_histogram()+
  facet_wrap(~CURREL)
ggplot(data = data_subset2, aes(x=SPIRPER))+
  geom_histogram()+
  facet_wrap(~CURREL)
ggplot(data = data_subset3, aes(x=SPIRPER))+
  geom_histogram()+
  facet_wrap(~CURREL)
```


The spread for religiosity is shifted more to the higher (less religious) values than spirituality: there were only two religions where the plurality (largest spike) was 1 (most religious).

```{r, message=FALSE}
data$RELPER[data$RELPER == 99] <- 5

data_subset1 <- data |>
  filter(CURREL %in% c("Protestant", "Catholic", "Unaffiliated"))
data_subset2 <- data |>
  filter(CURREL %in% c("Mormon", "Jewish", "Other Faith"))
data_subset3 <- data |>
  filter(CURREL %in% c("Orthodox", "JW", "Other Christian", "Muslim", "Buddhist", "Hindu", "Other Religion",  "Uninterpretable"))

ggplot(data = data_subset1, aes(x=RELPER))+
  geom_histogram()+
  facet_wrap(~CURREL)
ggplot(data = data_subset2, aes(x=RELPER))+
  geom_histogram()+
  facet_wrap(~CURREL)
ggplot(data = data_subset3, aes(x=RELPER))+
  geom_histogram()+
  facet_wrap(~CURREL)
```

# Statistical Analysis

To formally investigate the relationships suggested by the exploratory data analysis, we apply several statistical methods, including analysis of variance (ANOVA), correlation analysis, and multiple linear regression. These methods allow us to test for group differences, measure associations between variables, and identify key predictors of religiosity.

## Section 1: Group Differences

We begin by examining whether religiosity differs across religious traditions.

Hypothesis Test:
H0: No difference in religiosity between groups.
Ha: There is a difference.

```{r}
data_filtered <- subset(data, RELPER != 5)

anova_model <- aov(RELPER ~ CURREL, data = data_filtered)
summary(anova_model)

data_filtered %>%
  group_by(CURREL) %>%
  summarise(mean_religiosity = mean(RELPER, na.rm = TRUE),
            n = n()) %>%
   arrange(mean_religiosity)
```

Interpret the what we get from anova model result....

Check Assumptions for ANOVA:
1. Independence
2. Normality within groups
3. Equal variance

```{r}
plot(anova_model)
```


... Interpret assumptions.

## Section 2: Spirituality vs Religiosity

Next, we examine the relationship between spirituality and religiosity, as the exploratory analysis suggested that even unaffiliated individuals report relatively high levels of spirituality.

```{r}
cor(data$RELPER, data$SPIRPER, use="complete.obs")

model1 <- lm(RELPER ~ SPIRPER, data=data)
summary(model1)
```
Choose a plot to see the relationship:

```{r}
ggplot(data = data) +
  geom_jitter(aes(x = SPIRPER, y = RELPER), width = 0.2, height = 0.2, alpha = 0.3) +
  geom_smooth(method = "lm", aes(x = SPIRPER, y = RELPER), color = 'green')

ggplot(data, aes(x = factor(SPIRPER), y = RELPER)) +
  geom_boxplot()

data |> 
  group_by(SPIRPER) |> 
  summarize(mean_rel = mean(RELPER, na.rm=TRUE)) |> 
  ggplot(aes(x = SPIRPER, y = mean_rel)) +
  geom_point() +
  geom_line()
```

Interpret....

## Section 3: Behavioral Predictors

While spirituality captures internal beliefs, religiosity may also be influenced by observable behaviors such as attendance and prayer. We therefore examine whether these behavioral variables are significant predictors of religiosity.

```{r}
no99 <- subset(data, RELPER != 5 & ATTNDPERRLS != 99 & PRAY != 99 & PRAC_B != 99 & PRAC_D != 99)

model2 <- lm(RELPER ~ ATTNDPERRLS + PRAY + PRAC_B + PRAC_D, data=no99)
summary(model2)

protestant_data <- subset(no99, CURREL == "Protestant")
model_protestant <- lm(RELPER ~ ATTNDPERRLS + PRAY + PRAC_B + PRAC_D,
                       data = protestant_data)
summary(model_protestant)

catholic_data <- subset(no99, CURREL == "Catholic")
model_catholic <- lm(RELPER ~ ATTNDPERRLS + PRAY + PRAC_B + PRAC_D,
                       data = catholic_data)
summary(model_catholic)

muslim_data <- subset(no99, CURREL == "Muslim")
model_muslim <- lm(RELPER ~ ATTNDPERRLS + PRAY + PRAC_B + PRAC_D,
                       data = muslim_data)
summary(model_muslim)
```
Hypothesis Testing:
H0:
Ha:

Interpret the result...

Check for Assumptions of Regression:
1. Linearity
2. Nearly normal residuals
3. Constant variability

## Section 4: Political Predictors of Religiosity

```{r}
data_clean <- subset(data, RELPER != 99 & ABRTLGL != 99 & GAYMARR != 99 & QB2A != 99 & CHNG_C)

model3 <- lm(RELPER ~ ABRTLGL + GAYMARR + QB2A + CHNG_C, data = data_clean)
summary(model3)

data_clean_spir <- subset(data, SPIRPER != 99 & ABRTLGL != 99 & GAYMARR != 99 & QB2A != 99 & CHNG_C)

data_spir <- lm(SPIRPER ~ ABRTLGL + GAYMARR + QB2A + CHNG_C, data = data_clean)
summary(data_spir)


```
It would appear that removal of the 99s which represent those who did not answer made a substantial difference on the fit of the model. 
Furthermore, support for gay marriage predicts religiosity well, as does support for transgender individuals. However, acceptance of homosexuality is an incredibly poor predictor of religiosity in the body public. 

Alternatively, support for abortion, homosexuality, and gay marriage predicts _spirituality_ but support for transgender individuals does not predict _spirituality_

```{r}
data_clean2 <- subset(data, RELPER != 99 & ABRTLGL != 99 & GAYMARR != 99 & QB2A != 99 & CHNG_C != 99)


catholic_pol <- data_clean2 |> filter(CURREL == "Catholic")
catholic_polly <- lm(RELPER ~ ABRTLGL + GAYMARR + QB2A + CHNG_C, data = catholic_pol)
summary(catholic_polly)


protestant_pol <- data_clean2 |> filter(CURREL == "Protestant")
protestant_polly <- lm(RELPER ~ ABRTLGL + GAYMARR + QB2A + CHNG_C, data = protestant_pol)
summary(protestant_polly)

unaffiliated_pol <- data_clean2 |> filter(CURREL == "Unaffiliated")
unaffiliated_polly <- lm(RELPER ~ ABRTLGL + GAYMARR + QB2A + CHNG_C, data = unaffiliated_pol)
summary(unaffiliated_polly)


```


While support for abortion, same sex marriage, and homosexuality is a statistically significant predictor of religiousity in Catholics and people who are unaffiliated, support for these causes is not statistically significant in predicting religiosity in Protestants. Interstingly, greater acceptance of transgender individuals cannot be used to predict religiosity in people who are unaffiliated. 

```{r}
data_clean3 <- subset(data, RELPER != 99 & CHNG_A != 99 & CHNG_B != 99 & DIVRELPOP != 99 & DIVRACPOP != 99 & QB2C != 99)

catholic_pol2 <- data_clean3 |> filter(CURREL == "Catholic")
catholic_polly2 <- lm(RELPER ~ CHNG_A + CHNG_B + DIVRELPOP + DIVRACPOP + QB2C, data = catholic_pol)
summary(catholic_polly2)

protestant_pol2 <- data_clean3 |> filter(CURREL == "Protestant")
protestant_polly2 <- lm(RELPER ~ CHNG_A + CHNG_B + DIVRELPOP + DIVRACPOP + QB2C, data = protestant_pol2)
summary(protestant_polly2)

unaffiliated_pol2 <- data_clean3 |> filter(CURREL == "Unaffiliated")
unaffiliated_polly2 <- lm(RELPER ~ ABRTLGL + GAYMARR + QB2A + CHNG_C, data = unaffiliated_pol2)
summary(unaffiliated_polly2)

vars_to_plot <- c("RELPER", "CHNG_A", "CHNG_B", "DIVRELPOP", "DIVRACPOP", "QB2C")

ggpairs(data_clean3, 
        columns = vars_to_plot,
        title = "Correlation Matrix for RELPER Predictors",
        upper = list(continuous = wrap("cor", size = 4)),    # Shows correlation coefficients
        lower = list(continuous = "smooth_loess"),           # Adds a trend line to scatter plots
        diag = list(continuous = "densityDiag")              # Shows the distribution of each variable
) +
theme_minimal()

```

Support for a growing immigrant population is not predictive of religiousity, but 

## Full Model, Search for the Best Model

Finally, we construct a comprehensive model that incorporates spirituality, behavioral measures, belief variables, and religious affiliation to determine which factors are most important in predicting religiosity.

```{r}
final_model <- lm(RELPER ~ SPIRPER + ATTNDPERRLS + PRAY + PRAC_B + PRAC_D + GOD + HVN + HLL + CURREL, data=data)
summary(final_model)
```

Check for Assumptions of Regression:
1. Linearity
2. Nearly normal residuals
3. Constant variability

Dropping the variables that has p-value higher than 0.05. After dropping will model be better?
