---
title: "religion_project"
author: "Rachel Chen, Topher Hanagan, Jakob Hildebrandt"
date: "2026-03-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(readr)
library(dplyr)
# import data
data <- read_csv("Test.csv")

data <- data |> select(-REGION, -DENOMREC2, -FAMILY, -PROTFAM, -CURREL2, -FRMREL2, -SPREL, -SPREL2, -SPBORNFINAL, -KIDS4ANDUNDERREC, -KIDS5TO12REC, -KIDS13TO17REC, -CITIZEN, -BRANCH, -BIRTHHALFDECADE, -RESPONDENT_BIRTHREGION, -MOTHER_BIRTHREGION, -FATHER_BIRTHREGION, -YEARSINUSREC, -ORIENTMOD, -STATE, -MSA)
```

This dataset has information about over 36,000 individuals, with columns including information about the state or origin, happiness, questions about attitudes about other religions and immigrants, religious teachings, perceived spirituality/religiosity, marriage status, political leaning, family income, among others. This data comes from the Pew Research Center, a reputable source for surveys on the American population.

### Ideas and Questions:

Some potential areas we could explore with this data include seeing which variables are most associated with religiosity.

1) Test people's faithfulness as predictors or stand-ins. How important is religion in your life? How important is this to you when making decisions about right and wrong?

2) Are you personally a member of a church, synagogue, mosque, temple, or other house of worship?

3) Look at people who consider themselves religious vs actually attend religious services. 

4) Try to find the correlation between consider yourself religious and attendance. Could also use a composite for praying and attendance to as a stand-in for "religiosity"

5) Check how the religion related with family or parents. How parents religion affect your religion? Is father religion or mother religion same as your?

6) How religious based of age(birth)? Try to find out did the state affect how you religious?

7) How marriage affect the religion?

8) How support for abortion vs gay marriage as a predictor for party affiliation?


# Driving Question:
How well can we predict religiosity based on factors including church attendance, family relationships, marriage, political opinions.